Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2024-12-20 manuscript.Rmd
restructure manuscript with new model and elevation gradient

In this study, we aim to better understand traditional AFS in Timor-Leste by quantifying the carbon storage in both vegetation biomass and soil of five traditional AFS: cropping systems with fallow, silvopastures, young agroforests, home gardens, and forest gardens. We hypothesize that there are measurable differences in carbon in aboveground vegetation biomass (hereinafter referred to as above ground carbon or AGC; Hypothesis 1a) and SOC (Hypothesis 1b) stocks between these different types of traditional AFS. In line with @Ma2020, we anticipate that there is a positive relationship between AGC and SOC stocks in the topsoil layer (0-10 cm; Hypothesis 2a), but no relationship with SOC below 10 cm (Hypothesis 2b).

@@

# get coordinates
data_coord <- data_coord[, .(long = mean(as.numeric(LONG)), lat = mean(as.numeric(LAT))), .(N_PLOT)]

@@

```{r get-env-data}
# precipitation from CHELSA
if (!file.exists("data/CHELSA_bio12_1981-2010_V.2.1.tif")) {
  url <- "https://os.zhdk.cloud.switch.ch/envicloud/chelsa/chelsa_V2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio12_1981-2010_V.2.1.tif"
  destfile <- "data/CHELSA_bio12_1981-2010_V.2.1.tif"
  download.file(url, destfile, method = "curl")

@@

# 30m resolution SRTM
if (!file.exists("data/CHELSA_bio12_1981-2010_V.2.1.tif")) {
  url <- "https://srtm.csi.cgiar.org/wp-content/uploads/files/srtm_5x5/TIFF/srtm_62_14.zip"
  destfile <- "data/srtm_62_14.zip"
  download.file(url, destfile, method = "curl")
  unzip(destfile, exdir = "data", overwrite = TRUE)
}

@@

```{r extract-env-data}
# extract elevation and rainfall for the 2 sites
site_env = data_plot[, as.list(expand.grid(seq(min(long), max(long), 0.001), 
                                           seq(min(lat), max(lat), 0.001))),
                     .(site)]

@@

Briefly, the CF are plantations of a few non-perennial crops (e.g., maize [*Zea mays*], sweet potatoes [*Ipomoea batatas*], groundnuts [*Arachis hypogaea*], and squash gourd [*Cucurbita moschata*]) which are left fallow after one to three years of cropping. SP are vast areas (generally 200 to 500 ha) where cattle (*Bos taurus*, *Bos domesticus*, *Bubalus bubalis*) are grazed under the shade of palms, like coconut (*Cocos nucifera*), and trees, like tamarind (*Tamarindus indica*) or eucalyptus (*Eucalyptus alba*) and jujube (*Ziziphus sp.*). YA are agricultural plots where farmers plant rows of trees, like teak (*Tectona grandis*) and mahogany (*Swietenia mahagoni*), spaced approximately 5 x 10 m apart, which are maintained for two to ten years and then either cut for wood or left to grow to become HG or FG (see Figure S\@ref(fig:seq-diagram)). HG are plots of land located close to houses where crops and livestock (poultry [*Gallus gallus domesticus*], pigs [*Sus domesticus*], cattle, etc.) are kept in the shade of numerous trees, like Breadfruit (*Artocarpus altilis*), candlenut (*Aleurites moluccanus*) and palms, like areca palm (*Areca catechu*), for multiple uses (wood, fruit, fiber, etc.). FG are generally former HG (over 50 years old) which have been abandoned, for example as a result of the war, and are now used mainly for gathering fruits (e.g. mango [*Mangifera indica*]) and non-timber forest products (e.g. candlenut) with limited human influence.

@@

### Tree measurements

@@

### Soil measurements

In each of the 30 sample plots, a composite of three samples was taken from three of the five 10 x 10 m subplots, following a transect covering the intra-plot variability and at a minimum distance of 1 m from any tree (Figure S\@ref(fig:design)). Soils were sampled with a 15 cm auger at 0-10 cm and 20-30 cm depth; due to the size of the auger we could not accurately sample the intermediate soil layer. Soils samples were then air-dried and sieved at 2 mm. They underwent standard physical and chemical analyses at Cirad laboratories in Montpellier, France (Table S\@ref(tab:site-soil)). Total organic carbon was determined by dry combustion and measured with a thermal conductive detector.

In addition, at one location at the center of the plot, soils were sampled with a cylinder of 250 cm^3^ at depths of 2-7 cm, 12-17 cm and 22-27 cm. Each sample was dried at 105°C in a classic oven until a stable dry mass was reached and then sieved to 2 mm. The fine fraction (\< 2 mm) was weighed to be used in the SOC stocks calculation (equation \@ref(eq:soc)).

@@

$$SOC_p = \frac{\sum_i Corg_i \cdot (mfine_i \cdot 10^{-6}) \cdot (Splot \cdot \Delta d)}{Vcyl}$$

Where $Corg_i$ is the SOC content at depth $i$ (either 0-10, 10-20 or 20-30 cm); $mfine_i$ is the mass of the fine fraction of the soil in the sampling cylinder at depth $i$ (in g), converted into Mg with a factor of 10$^{-6}$; $Splot$ is the surface of a 1-ha plot (10$^8$ cm$^2$); $\Delta d$ is the difference in sample depths (= 10 cm) and $Vcyl$ is the volume of the sampling cylinder (250 cm$^3$). This calculation corresponds to the method recommended by @Poeplau2017 (M4), which correctly takes into account the coarse fraction of the soil. At depth $i$ = 10-20 cm, where SOC content was not measured (see "Soil measurements" section), we estimated SOC content as a function of SOC content at 0-10 cm and 20-30 cm, assuming a negative exponential model of SOC content decrease with soil depth [@Mishra2009; @Murphy2019]:

\begin{equation}
Corg_{10,20} = Corg_{0,10} \cdot \sqrt{\frac{Corg_{20,30}}{Corg_{0,10}}} (\#eq:intersoc)
\end{equation}

@@

```{r add-other-soil-properties}
data_soil_summ <- data_soil |> 
  subset(!is.na(TC)) |> 
  melt(id.vars = c("N_PLOT", "SAMPLE_DEPTH"), 
       measure.vars= c("CLAY", "FINE SILT", "pH\n", "Ca", "CEC\n"))  |> 
  setDT()
data_soil_summ[, variable := gsub("\n", "", variable)]
data_soil_summ <- 
  data_soil_summ[, .(value = mean(as.numeric(value), na.rm = TRUE)), .(N_PLOT, variable)] |> 
  dcast(N_PLOT ~ variable)
data_soil_summ[, fine_soil := CLAY + `FINE SILT`] 
data_sim <- merge(data_sim, data_soil_summ)  
```

@@

We calibrated the following models for AGC stocks of plot $i$ in site $k$ and AFS type $j$:

$$AGC_{i,j,k}  = \mu_A + \alpha_{A,j} + \beta_{A,k} + \epsilon_{A,i}$$
with $\mu_A$ the average AGC stocks in our study area, $\alpha_{A,j}$ and  $\beta_{A,k}$ the random AFS and site effects, respectively distributed as:

$$\alpha_{A,j} \sim \mathcal{N}(0,\sigma_{\alpha A}^2)$$

$$\beta_{A,k} \sim \mathcal{N}(0,\sigma_{\beta A}^2)$$
And $\epsilon_{A,i}$ the random error, distributed as:

$$\epsilon_{A,i} \sim \mathcal{N}(0,\sigma_A^2)$$
$\sigma_A$, $\sigma_{\alpha A}$, and $\sigma_{\beta A}$ are the standard deviation of the random error, the AFS effect and the site effect respectively.

Similarly, we modelled SOC stocks of plot $i$ in site $k$ and AFS type $j$ as:

$$SOC_{i,j,k} = \mu_S +  \alpha_{Sj} + \beta_{Sk} + \theta_1 agc_i + \theta_2 fine_i+ \epsilon_{S,i}$$
with $\mu_S$ the average SOC stocks in our study area; $\theta_1$ the effect of AGC stocks on SOC stocks and $agc_i$ the standardised AGC stocks of plot $i$; $\theta_2$ the effect of fine soil fraction (clay + fine silt) on SOC stocks and $fine_i$ the standardised fine soil fraction of plot $i$. We standardised AGC stocks and fine soil fraction by scaling and centring values.

$\alpha_{S,j}$ and $\beta_{S,k}$ are the random AFS and site effects, respectively distributed as:

$$\alpha_{S,j} \sim \mathcal{N}(0,\sigma_{\alpha S}^2)$$

$$\beta_{S,k} \sim \mathcal{N}(0,\sigma_{\beta S}^2)$$
And $\epsilon_{S,i}$ the random error, distributed as:

$$\epsilon_{S,i} \sim \mathcal{N}(0,\sigma_S^2)$$
$\sigma_S$, $\sigma_{\alpha S}$, and $\sigma_{\beta S}$ are the standard deviation of the random error, the AFS effect and the site effect respectively.

Calibration was carried out using an adapted form of the Hamiltonian Monte Carlo using Stan’s programming language [@Carpenter2017]. We used the standard half-normal distribution as the prior for all standard deviation parameters, and the standard normal distribution for both $\theta_1$ and $\theta_2$ parameters. To propagate uncertainties on the model outputs, we calibrated the model for each of the 1000 iterations and sampled one set of parameter values in the posterior distributions for each iteration.

```{r}
stan_model <- "
data {
  int<lower=0> N;
  int<lower=1> AF;
  int<lower=1> S;
  int<lower=1,upper=AF> afs[N];
  int<lower=1,upper=S> site[N];
  real clay_sc[N];
  real agc_sc[N];
  real<lower=0> agc[N];
  real<lower=0> soc[N];
  real<lower=0> soc10[N];
}

parameters {
  real mu_a;
  real delta_afs_a[AF];
  real delta_site_a[S];
  real<lower=0> sigma_a;

  real mu_s;
  real delta_afs_s[AF];
  real delta_site_s[S];
  real theta_agc_s;
  real theta_clay_s;
  real<lower=0> sigma_s;

  real mu_s10;
  real delta_afs_s10[AF];
  real delta_site_s10[S];
  real theta_agc_s10;
  real theta_clay_s10;
  real<lower=0> sigma_s10;

@@

model {
  real pred_a[N];
  real pred_s[N];
  real pred_s10[N];

  for (i in 1:N) {
    pred_a[i] = mu_a + delta_afs_a[afs[i]] + delta_site_a[site[i]];
    pred_s[i] = mu_s + theta_agc_s*agc_sc[i] + theta_clay_s*clay_sc[i] + delta_afs_s[afs[i]] + delta_site_s[site[i]];
    pred_s10[i] = mu_s10 + theta_agc_s10*agc_sc[i] + theta_clay_s10*clay_sc[i] + delta_afs_s10[afs[i]] + delta_site_s10[site[i]];
  }

  agc ~ normal(pred_a, sigma_a);
  soc ~ normal(pred_s, sigma_s);
  soc10 ~ normal(pred_s10, sigma_s);

@@

  //  priors
  theta_agc_s ~ normal(0,1);
  theta_agc_s10 ~ normal(0,1);
  theta_clay_s ~ normal(0,1);
  theta_clay_s10 ~ normal(0,1);
  sigma_a ~ normal(0,1);
  sigma_s ~ normal(0,1);
  sigma_s10 ~ normal(0,1);
  sigma_site_a ~ normal(0,1);
  sigma_afs_a ~ normal(0,1);
  sigma_site_s ~ normal(0,1);
  sigma_afs_s ~ normal(0,1);
  sigma_site_s10 ~ normal(0,1);
  sigma_afs_s10 ~ normal(0,1);
}

@@

if (!file.exists("outputs/data_par.csv")) {
  library(parallel)

  cl <- makeCluster(detectCores()-1)
  clusterEvalQ(cl, {library(rstan); library(data.table)})
  clusterExport(cl, "data_sim"); clusterEvalQ(cl, data_sim)
  clusterExport(cl, "stan_model"); clusterEvalQ(cl, stan_model)

  data_par <- parLapply(cl, unique(data_sim$iter), function(i) {
    rstan_options(auto_write = TRUE)

    data_stan <- data_sim |> 
      subset(iter==i) |> 
      dcast(N_PLOT + site + ID_AF + fine_soil ~ variable) |> 
      with(
        list(
          N = length(N_PLOT),
          AF = length(unique(ID_AF)),
          S = length(unique(site)),
          afs = as.numeric(as.factor(ID_AF)), 
          site = as.numeric(as.factor(site)), 
          clay_sc = (fine_soil-mean(fine_soil))/sd(fine_soil),
          agc_sc = (AGC-mean(AGC))/sd(AGC),
          agc = AGC,
          soc = SOC,
          soc10 = SOC10
        ))

    fit <- stan(model_code = stan_model, data = data_stan, 
                iter = 60, warmup = 50, chains = 1)
    pars <- extract(fit) |> 
      lapply(function(x) {
        df <- data.table(iter_stan = 1:dim(x)[1], x) |> 
          melt(id.vars = "iter_stan", variable.name = "level_num")  
        return(df)
      }) |> 
      rbindlist(idcol = "param")
    return(pars)
  }) |> rbindlist(idcol = "iter")

  write.csv(data_par, file = "outputs/data_par.csv", row.names = FALSE)

  stopCluster(cl)

} else data_par <- read.csv("outputs/data_par.csv") |> setDT()

data_par[, effect := tstrsplit(param, "_")[[2]]]
data_par[! effect %in% c("afs", "site", "agc", "clay"), effect := NA]
data_par[, stock := tstrsplit(param, "_")[[3]]]
data_par[is.na(stock) & param != "lp__", stock := tstrsplit(param, "_")[[2]]]
data_par[stock == "a", stock := "agc"]
data_par[stock == "s", stock := "soc"]
data_par[stock == "s10", stock := "soc10"]
data_par[, param := tstrsplit(param, "_")[[1]]]

data_par[, level_num := as.numeric(gsub("V", "", level_num))]
data_par[effect == "afs" & !is.na(level_num), 
         level := levels(as.factor(data_sim$ID_AF))[level_num]]
data_par[effect == "site" & !is.na(level_num),
         level := levels(as.factor(data_sim$site))[level_num]]

df_mu_delta <- 
  data_par[param == "delta" & !is.na(effect)] |> 
  dcast(iter + iter_stan + stock + effect + level ~ param) |> 
  merge(dcast(data_par[param == "mu"], iter + iter_stan + stock ~ param))
df_mu_delta[, `:=`(value = mu + delta, param = "pred_med")]
data_par <- rbind(
  data_par[, c("param", "iter", "iter_stan", "stock", "effect", "level", "value")], 
  df_mu_delta[, c("param", "iter", "iter_stan",  "stock", "effect", "level", "value")]
)

@@

```{r group-comparison}
# shapiro wilk test - normality
dataC_plot[variable %in% c("AGC", "SOC") & plot != 6, .(
  shap = shapiro.test(log(mean))$p.value
), .(afs, variable)]
# levene test of homoscedasticity
dataC_plot[variable %in% c("AGC", "SOC") & plot != 6, 
           leveneTest(log(mean) ~ afs*site)$`Pr(>F)`[1], .(variable)]

# 2-way anova 
anova_table = 
  dataC_plot[variable %in% c("AGC", "SOC") & plot != 6, 
             .(pvalue = round(anova(lm(log(mean) ~ site + afs))$`Pr(>F)`[1:2], 3)), 
             .(variable)]
anova_table$factor = rep(c("site", "afs"), 2)
```

```{r lmereg}
data_reg = melt(dataC_plot[plot != 6], measure.vars = c("mean", "lwr", "upr"), variable.name = "stat") |> 
  dcast(plot + site+ afs ~ variable + stat)

## mixed effect linear regression
lmeReg <- lmer(log(SOC_mean) ~ log(AGC_mean) + (1|site), data = data_reg)
# Nakagawa’s R2 for mixed models
R2 = r2_nakagawa(lmeReg)
summary(lmeReg)

# same for topsoil layer (0-10 cm)
lmeReg10 <- lmer(log(SOC10_mean) ~ log(AGC_mean) + (1|site), data = data_reg)
# Nakagawa’s R2 for mixed models
R210 = r2_nakagawa(lmeReg10)
summary(lmeReg10)

dfreg = 
  data.table(
    depth = c("Soil depth: 0-30 cm", "Soil depth: 0-10 cm"), 
    intercept = c(summary(lmeReg)$coefficients[1,1], summary(lmeReg10)$coefficients[1,1]), 
    slope = c(summary(lmeReg)$coefficients[2,1], summary(lmeReg10)$coefficients[2,1]), 
    pvalue = c(summary(lmeReg)$coefficients[2,5], summary(lmeReg10)$coefficients[2,5]),
    R2_cond = c(R2$R2_conditional, R210$R2_conditional), 
    R2_marg = c(R2$R2_marginal, R210$R2_marginal)
  )
```

@@

## Total carbon stocks

@@

## Aboveground carbon

>>>>

[paragraph on possible drivers]
C stocks (in biomass and soil) of AFS could be driven by a number of factors: 
- agricultural practices (ref)
- climate (ref)
- soil properties (especially for SOC) > in particular, fine content

In this study, we aim to better understand traditional AFS in Timor-Leste by quantifying the carbon storage in both vegetation biomass and soil of five traditional AFS: cropping systems with fallow, silvopastures, young agroforests, home gardens, and forest gardens. [xxx 2 sites] We hypothesize that there are measurable differences in carbon in aboveground vegetation biomass (hereinafter referred to as above ground carbon or AGC; Hypothesis 1a) and SOC (Hypothesis 1b) stocks between these different types of traditional AFS. In line with @Ma2020, we anticipate that there is a positive relationship between AGC and SOC stocks in the topsoil layer (0-10 cm; Hypothesis 2a), but no relationship with SOC below 10 cm (Hypothesis 2b).

@@

# get coordinates
data_coord <- data_coord[, .(
  long = mean(as.numeric(LONG)), 
  lat = mean(as.numeric(LAT))
), .(N_PLOT)]

@@

```{r get-env-data}
# temperature and precipitation from CHELSA
if (!file.exists("data/CHELSA_bio1_1981-2010_V.2.1.tif")) {
  url <- "https://os.zhdk.cloud.switch.ch/chelsav2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio1_1981-2010_V.2.1.tif"
  destfile <- "data/CHELSA_bio1_1981-2010_V.2.1.tif"
  download.file(url, destfile, method = "curl")

@@

if (!file.exists("data/CHELSA_bio12_1981-2010_V.2.1.tif")) {
  url <- "https://os.zhdk.cloud.switch.ch/chelsav2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio12_1981-2010_V.2.1.tif"
  destfile <- "data/CHELSA_bio12_1981-2010_V.2.1.tif"
  download.file(url, destfile, method = "curl")

# 30m resolution SRTM
if (!file.exists("data/srtm_62_14.zip")) {
  url <- "https://srtm.csi.cgiar.org/wp-content/uploads/files/srtm_5x5/TIFF/srtm_62_14.zip"
  destfile <- "data/srtm_62_14.zip"
  download.file(url, destfile, method = "curl")
  unzip(destfile, exdir = "data", overwrite = TRUE)
}

@@

```{r extract-env-data}
# extract elevation, precipitation and temperature for all plots
data_plot[, `:=`(
  elev = terra::extract(terra::rast("data/srtm_62_14.tif"), 
                        cbind(long, lat))[[1]], 
  prec = terra::extract(terra::rast("data/CHELSA_bio12_1981-2010_V.2.1.tif"), 
                        cbind(long, lat))[[1]], 
  temp = terra::extract(terra::rast("data/CHELSA_bio1_1981-2010_V.2.1.tif"), 
                        cbind(long, lat))[[1]]
)]

cor(data_plot[, c("elev", "prec", "temp")])

# extract elevation and rainfall for the 2 sites
site_env = data_plot[, as.list(expand.grid(seq(min(long), max(long), 0.001), 
                                           seq(min(lat), max(lat), 0.001))),
                     .(site)]

@@

Briefly, the CF are plantations of a few non-perennial crops (e.g., maize [*Zea mays*], sweet potatoes [*Ipomoea batatas*], groundnuts [*Arachis hypogaea*], and squash gourd [*Cucurbita moschata*]) which are left fallow after one to three years of cropping. SP are vast areas (generally 200 to 500 ha) where cattle (*Bos taurus*, *Bos domesticus*, *Bubalus bubalis*) are grazed under the shade of palms, like coconut (*Cocos nucifera*), and trees, like tamarind (*Tamarindus indica*) or eucalyptus (*Eucalyptus alba*) and jujube (*Ziziphus sp.*). YA are agricultural plots where farmers plant rows of trees, like teak (*Tectona grandis*) and mahogany (*Swietenia mahagoni*), spaced approximately 5 x 10 m apart, which are generally left to grow to become HG or FG (see Figure S\@ref(fig:seq-diagram)). HG are plots of land located close to houses where crops and livestock (poultry [*Gallus gallus domesticus*], pigs [*Sus domesticus*], cattle, etc.) are kept in the shade of numerous trees, like Breadfruit (*Artocarpus altilis*), candlenut (*Aleurites moluccanus*) and palms, like areca palm (*Areca catechu*), for multiple uses (wood, fruit, fiber, etc.). FG are generally former HG (over 50 years old) which have been abandoned, for example as a result of the war, and are now used mainly for gathering fruits (e.g. mango [*Mangifera indica*]) and non-timber forest products (e.g. candlenut) with limited human influence.

@@

@@

In each of the 30 sample plots, a composite of three samples was taken from three of the five 10 x 10 m subplots (including the plot center and two opposite corners; Figure S\@ref(fig:design)), covering the intra-plot variability and at a minimum distance of 1 m from any tree. Soils were sampled with a soil auger ( Model Bor Tanah - Dormer Standard Soil Auger, Adiguna Karya Persada, Bogor, Indonesia) at 0-10 cm and 20-30 cm depth,. As the length of the head auger (15 cm) was longer than our target soil depths (10 cm), we could not accurately sample the 10-20 cm soil layer. Soils samples were then air-dried and sieved at 2 mm. They underwent standard physical and chemical analyses at Cirad laboratories in Montpellier, France (Table S\@ref(tab:site-soil)). Total carbon content was determined by dry combustion, following the ISO 10694:1995 standard method, and measured with a thermal conductive detector (EA 1112 Elemental Analyser, Thermo Fischer Scientific, Netherlands). Carbonate content was determined with a calcimeter, following the ISO 10693 standard method. Soil organic carbon content was calculated as the difference of the total carbon and the carbonate contents.

In addition, at one location in the center of the plot, three 250 cm$^3$ soil samples were taken with a soil sampling cylinder (5 cm in height, 7.98 cm inner diameter, and 8.4 cm outer diameter; Eijkelkamp Soil & Water, Giesbeek, Netherlands)  at depths of 2-7 cm, 12-17 cm and 22-27 cm.  These depths were chosen to take account of the height of the cylinder (i.e. 5 cm), the destruction of the soil below the cylinder during its extraction, and to be associated with depths 0-10 cm, 10-20 cm and 20-30 cm, the depths at which the organic carbon content was calculated or estimated. Each sample was dried at 105°C in a classic oven until a stable dry mass was reached and then sieved to 2 mm. The fine fraction (\< 2 mm) was weighed to be used in the SOC stocks calculation (equation \@ref(eq:soc)).

@@

Where $Corg_i$ is the SOC content at depth i (either 0-10, 10-20 or 20-30 cm) as measured in the laboratory from the soil sampled by the auger and estimated at depth 10-20 cm (see details below); $mfine_i$ is the mass of the fine fraction of the soil in the sampling cylinder at depth i (in g), converted into Mg with a factor of 10$^-6$; $Surf\ =\ 10^8$ is the surface conversion factor from cm$^2$ to ha; $\Delta d$ is the difference in sample depths (= 10 cm) and $Vcyl$ is the volume of the sampling cylinder (250 cm$^3$). This calculation corresponds to the method M4 recommended by @Poeplau2017, which correctly considers the coarse fraction of the soil. We considered that $mfine_i$ at depths of 2-7 cm, 12-17 cm and 22-27 cm was associated with $Corg_i$ at depths of 0-10, 10-20 cm and 20-30 cm, respectively. At depth $i$ = 10-20 cm, where the carbon content was not measured (see “Soil measurements” section), we estimated SOC content as a function of SOC content at 0-10 cm and 20-30 cm, assuming a standard negative exponential model of SOC content decrease with soil depth [@Mishra2009;@Murphy2019]:

\begin{equation}
Corg_{10,20}=Corg_{0,10}\cdot\sqrt{\frac{Corg_{20,30}}{Corg_{0,10}}}
(\#eq:soc)
\end{equation}

@@

```{r add-other-soil-properties}
data_soil_summ <- data_soil |> 
  subset(!is.na(TC)) |> 
  melt(id.vars = c("N_PLOT", "SAMPLE_DEPTH"), 
       measure.vars= c("CLAY", "FINE SILT", "pH\n", "Ca", "CEC\n"))  |> 
  setDT()
data_soil_summ[, variable := gsub("\n", "", variable)]
data_soil_summ <- 
  data_soil_summ[, .(value = mean(as.numeric(value), na.rm = TRUE)), .(N_PLOT, variable)] |> 
  dcast(N_PLOT ~ variable)
data_soil_summ[, fine_soil := CLAY + `FINE SILT`] 
data_sim <- merge(data_sim, data_soil_summ)  
# add elevation
data_sim <- merge(data_sim, data_plot[, c("N_PLOT", "elev")])
```

@@

We modelled both AGC and SOC stocks as:

$$Y_{i,j,k} = \mu +  \theta \cdot X_i + \alpha_{j} + \beta_{k} + \epsilon_{i}$$
with $Y_{i,j,k}$ the stocks (either SOC or AGC) of plot $i$ in site $k$ and AFS type $j$

$\mu$ is the average stocks in our study area.

$\theta$ is the effect of the covariates on $Y$, and $X_i$ are the values of the covariates in plot $i$; they include elevation for AGC stocks [from NASA SRTM; @SRTM2024], or elevation [@SRTM2024], clay + fine soil content (from soil analysis; xxx methods) and AGC stocks for SOC stocks. All covariates were centred and scaled. Elevation was chosen as it was highly correlated (>95%) with both mean annual temperature and precipitation [@Karger2017]: it was thus a good indicator of a climate gradient in our dataset (see Supplementary information section xx). xxx justify these variables in the introduction

$\alpha_{j}$ and $\beta_{k}$ are the random AFS and site effects, respectively distributed as:

$$\alpha_{j} \sim \mathcal{N}(0,\sigma_{\alpha}^2)$$

$$\beta_{k} \sim \mathcal{N}(0,\sigma_{\beta}^2)$$
And $\epsilon_{i}$ the random error, distributed as:

$$\epsilon_{i} \sim \mathcal{N}(0,\sigma^2)$$
$\sigma$, $\sigma_{\alpha}$, and $\sigma_{\beta}$ are the standard deviation of the random error, the AFS effect and the site effect respectively.

Calibration was carried out using an adapted form of the Hamiltonian Monte Carlo using Stan’s programming language [@Carpenter2017]. We used the standard half-normal distribution as the prior for all standard deviation parameters, and the standard normal distribution for all $\theta$ parameters. To propagate uncertainties on the model outputs, we calibrated the model for each of the 1000 iterations and sampled one set of parameter values in the posterior distributions for each iteration.

```{r}
stan_model <- "
data {
  int<lower=0> N;
  int<lower=1> AF;
  int<lower=1> S;
  int<lower=1> K;
  int<lower=1,upper=AF> afs[N];
  int<lower=1,upper=S> site[N];
  vector[N] elev;
  matrix[N,K] covar;
  real<lower=0> agc[N];
  real<lower=0> soc[N];
  real<lower=0> soc10[N];
}

parameters {
  real mu_a;
  real delta_afs_a[AF];
  real delta_site_a[S];
  real theta_a;
  real<lower=0> sigma_a;

  real mu_s;
  real delta_afs_s[AF];
  real delta_site_s[S];
  vector[K] theta_s;
  real<lower=0> sigma_s;

  real mu_s10;
  real delta_afs_s10[AF];
  real delta_site_s10[S];
  vector[K] theta_s10;
  real<lower=0> sigma_s10;

@@

model {
  vector[N] pred_a;
  vector[N] pred_s;
  vector[N] pred_s10;

  // predictions
  for (i in 1:N) {
    pred_a[i] = mu_a + elev[i] * theta_a + delta_afs_a[afs[i]] + delta_site_a[site[i]];
    pred_s[i] = mu_s + dot_product(row(covar, i), theta_s) + delta_afs_s[afs[i]] + delta_site_s[site[i]];
    pred_s10[i] = mu_s10 + dot_product(row(covar, i), theta_s10) + delta_afs_s10[afs[i]] + delta_site_s10[site[i]];
  }

  // likelihoods
  agc ~ normal(pred_a, sigma_a);
  soc ~ normal(pred_s, sigma_s);
  soc10 ~ normal(pred_s10, sigma_s);

@@

  //  priors
  theta_s ~ normal(0,1);
  theta_s10 ~ normal(0,1);
  sigma_a ~ normal(0,1);
  sigma_s ~ normal(0,1);
  sigma_s10 ~ normal(0,1);
  sigma_site_a ~ normal(0,1);
  sigma_afs_a ~ normal(0,1);
  sigma_site_s ~ normal(0,1);
  sigma_afs_s ~ normal(0,1);
  sigma_site_s10 ~ normal(0,1);
  sigma_afs_s10 ~ normal(0,1);
}

@@

if (!file.exists("outputs/data_par.csv")) {
  library(rstan)
  rstan_options(auto_write = TRUE)

  pars_prev <- list.files("outputs", patter = "data_par_") |> 
    gsub(pattern = "data_par_|\\.csv", replacement = "") |> 
    as.numeric()

  for (n in setdiff(seq_len(100), pars_prev)) {
    data_par <- lapply(paste0("V", seq((n-1)*10+1, n*10)), function(i) {

      data_stan <- data_sim |> 
        subset(iter==i) |> 
        dcast(N_PLOT + site + ID_AF + fine_soil + elev ~ variable) |> 
        with(
          list(
            N = length(N_PLOT),
            AF = length(unique(ID_AF)),
            S = length(unique(site)),
            K = 3,
            afs = as.numeric(as.factor(ID_AF)), 
            site = as.numeric(as.factor(site)),
            elev = (elev-mean(elev))/sd(elev),
            covar = scale(cbind(elev, fine_soil, AGC)),
            agc = AGC,
            soc = SOC,
            soc10 = SOC10
          ))

      fit <- stan(model_code = stan_model, data = data_stan, 
                  iter = 60, warmup = 50, chains = 1)
      pars <- extract(fit) |> 
        lapply(function(x) {
          df <- data.table(iter_stan = 1:dim(x)[1], x) |> 
            melt(id.vars = "iter_stan", variable.name = "level_num")  
          return(df)
        }) |> 
        rbindlist(idcol = "param")
      return(pars[iter_stan==sample(iter_stan, 1), -"iter_stan"])
    }) |> rbindlist(idcol = "iter")

    write.csv(data_par, 
              file = paste0("outputs/data_par_", n, ".csv"), row.names = FALSE)
  }

  data_par <- list.files("outputs", pattern = "data_par_", full.names = TRUE) |> 
    lapply(read.csv) |> rbindlist(idcol = "iter10") 
  data_par[, `:=`(iter = (iter10-1)*10+iter, iter10 = NULL)]

  write.csv(data_par, "outputs/data_par.csv", row.names = FALSE)

  # list.files("outputs", pattern = "data_par_", full.names = TRUE) |>
  #   file.remove()
}

data_par <- read.csv("outputs/data_par.csv") |> setDT()

data_par[, effect := tstrsplit(param, "_")[[2]]]
data_par[! effect %in% c("afs", "site"), effect := NA]
data_par[, stock := tstrsplit(param, "_")[[3]]]
data_par[is.na(stock) & param != "lp__", stock := tstrsplit(param, "_")[[2]]]
data_par[stock == "a", stock := "agc"]
data_par[stock == "s", stock := "soc"]
data_par[stock == "s10", stock := "soc10"]
data_par[, param := tstrsplit(param, "_")[[1]]]

data_par[, level_num := as.numeric(gsub("V", "", level_num))]
data_par[effect == "afs" & !is.na(level_num), 
         level := levels(as.factor(data_sim$ID_AF))[level_num]]
data_par[effect == "site" & !is.na(level_num),
         level := levels(as.factor(data_sim$site))[level_num]]
data_par[param == "theta",
         level := c("elev", "fine", "agc")[level_num]]
data_par[param == "theta" & stock == "agc", level := "elev"]

# get levels of "delta" parameters
df_mu_delta <- 
  data_par[param == "delta" & !is.na(effect)] |> 
  dcast(iter + stock + effect + level ~ param) |> 
  merge(dcast(data_par[param == "mu"], iter + stock ~ param))
df_mu_delta[, `:=`(value = mu + delta, param = "pred_med")]
data_par <- rbind(
  data_par[, c("param", "iter", "stock", "effect", "level", "value")], 
  df_mu_delta[, c("param", "iter", "stock", "effect", "level", "value")]
)

@@

@@

AGC stocks ranged between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]` Mg C ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]` Mg C ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`% (Table \@ref(tab:summary); Figure \@ref(fig:barplot)). AGC stocks varied substantially between AFS, but not between sites (Figure \@ref(fig:effect-afs-site)). The AFS with the lowest AGC stocks were the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]` Mg C ha$^{-1}$; the AFS with the highest AGC stocks were the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]` Mg C ha$^{-1}$ (Table \@ref(tab:summary); Figure \@ref(fig:barplot)).

SOC stocks varied between `r dataC_plot[variable=="SOC", round(min(mean))]` and `r dataC_plot[variable=="SOC", round(max(mean))]` Mg C ha$^{-1}$, with an average of `r dataC_plot[variable=="SOC", round(mean(mean))]` Mg C ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="SOC", round(sd(mean)/mean(mean)*100)]`% (Table \@ref(tab:summary); Figure \@ref(fig:barplot)). SOC stocks varied substantially between sites, but not between AFS (Figure \@ref(fig:effect-afs-site)). The SOC stocks in Gariuai were on average higher (`r dataC_site[site=="Gariuai" & variable=="SOC", round(mean)]` Mg C ha$^{-1}$) than those in Osso-Luga (`r dataC_site[site=="Osso-Luga" & variable=="SOC", round(mean)]` Mg C ha$^{-1}$; Figure \@ref(fig:barplot)).

@@

## Effect of elevation and fine soil on AGC and SOC stocks

## Relationship between AGC and SOC stocks

The effect of AGC on SOC was close to zero (Figure \@ref(fig:plot-reg)).
